---
title: "Predictions for Driven Data"
authors: "Madison Hobbs and Jenn Havens"
date: "12/13/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
TestPred <- read.csv("./DengAI_Predicting_Disease_Spread_-_Test_Data_Features.csv")

load(file = "sj_model.rda",.GlobalEnv)
load(file = "iq_model.rda",.GlobalEnv)
```

We use our models for each city to predict the number of dengue cases in the data sets for which Driven Data wants predictions.

# Missing Values

First, we impute missing values using Imputation via Bagging (the same way we dealt with missing values in the training set).

```{r}
set.seed(6)
impTest <- preProcess(TestPred, "bagImpute")
test_noNA <- predict(impTest, TestPred)
```

# Create Time Lagged Variables

As in our training data, we generate time lagged weather variables from the data given.

```{r makeTimeLags}
lag_num_weeks = 12

#make on col which gives year and week of year info
tempData <- mutate(test_noNA, year_week_temp = paste(as.character(year), "_", as.character(weekofyear), sep="")) %>% mutate(lag_year_week_temp = paste(year(as.Date(week_start_date)-weeks(lag_num_weeks)), "_", week(as.Date(week_start_date)-weeks(lag_num_weeks)), sep=""))

#make year_week and lag_year_week of yyyy_ww format
tempData <- mutate(tempData, year_week = ifelse(nchar(year_week_temp) == 7, year_week_temp, sub("_", "_0", year_week_temp))) %>% mutate(lag_year_week = ifelse(nchar(lag_year_week_temp) == 7, lag_year_week_temp, sub("_", "_0", lag_year_week_temp)))

tempDataIQ <- filter(tempData, city == "iq")
tempDataSJ <- filter(tempData, city == "sj")

### San Juan Data for Model
#want lag for: precipitation_amt_mm and station_avg_temp_c for SJ
lag_week_temp <- c()
lag_week_precip <- c()
lag_week_humid <- c()
for (i in 1:length(tempDataSJ$lag_year_week)){
  if (sum(grepl(tempDataSJ$lag_year_week[i], tempDataSJ$year_week)) == 1){
    #find correct index for lag data and add that data to vectors
    lagWeeki <- grep(tempDataSJ$lag_year_week[i], tempDataSJ$year_week)
    lag_week_temp <- c(lag_week_temp, tempDataSJ$station_avg_temp_c[lagWeeki])
    lag_week_precip <- c(lag_week_precip, tempDataSJ$precipitation_amt_mm[lagWeeki])
    lag_week_humid <- c(lag_week_humid, tempDataSJ$reanalysis_relative_humidity_percent[lagWeeki])
  }else {
    lag_week_temp <- c(lag_week_temp, NA)
    lag_week_precip <- c(lag_week_precip, NA)
    lag_week_humid <- c(lag_week_humid, NA)
  }
}           

tempDataSJ$temp_lag <- lag_week_temp
tempDataSJ$precip_lag <- lag_week_precip
tempDataSJ$humidity_lag <- lag_week_humid 

# select only the predictors we want for the model
sj_data_for_model <- select(tempDataSJ, year, weekofyear, ndvi_ne, ndvi_nw, ndvi_se, ndvi_sw, precipitation_amt_mm, precip_lag, station_avg_temp_c, temp_lag, reanalysis_relative_humidity_percent, humidity_lag)


### Iquitos Data for Model
#want lag for: precipitation_amt_mm and station_avg_temp_c for IQ
lag_week_temp <- c()
lag_week_precip <- c()
lag_week_humid <- c()
for (i in 1:length(tempDataIQ$lag_year_week)){
  if (sum(grepl(tempDataIQ$lag_year_week[i], tempDataIQ$year_week)) == 1){
    #find correct index for lag data and add that data to vectors
    lagWeeki <- grep(tempDataIQ$lag_year_week[i], tempDataIQ$year_week)
    lag_week_temp <- c(lag_week_temp, tempDataIQ$station_avg_temp_c[lagWeeki])
    lag_week_precip <- c(lag_week_precip, tempDataIQ$precipitation_amt_mm[lagWeeki])
    lag_week_humid <- c(lag_week_humid, tempDataIQ$reanalysis_relative_humidity_percent[lagWeeki])
  }else {
    lag_week_temp <- c(lag_week_temp, NA)
    lag_week_precip <- c(lag_week_precip, NA)
    lag_week_humid <- c(lag_week_humid, NA)
  }
}           

tempDataIQ$temp_lag <- as.numeric(lag_week_temp)
tempDataIQ$precip_lag <- as.numeric(lag_week_precip)
tempDataIQ$humidity_lag <- as.numeric(lag_week_humid)

# select only the predictors we want for the model
iq_data_for_model <- select(tempDataIQ, year, weekofyear, ndvi_ne, ndvi_nw, ndvi_se, ndvi_sw, precipitation_amt_mm, precip_lag, station_avg_temp_c, temp_lag, reanalysis_relative_humidity_percent, humidity_lag)

iq_data_for_model <- filter(iq_data_for_model, humidity_lag != "NA")
sj_data_for_model <- filter(sj_data_for_model, humidity_lag != "NA")

# get good names for columns

iq_data_for_model <- iq_data_for_model %>% dplyr::rename(week_of_year = weekofyear) %>% dplyr::rename(relative_humidity_percent = reanalysis_relative_humidity_percent)

sj_data_for_model <- sj_data_for_model %>% dplyr::rename(week_of_year = weekofyear, relative_humidity_percent = reanalysis_relative_humidity_percent)
```

# Generate Predictions

We generate predictions and write them to csv in the format desired by Driven Data.

```{r}
sj_predicts <- predict(sj_model, sj_data_for_model)
iq_predicts <- predict(iq_model, iq_data_for_model)

sj_data_for_model <- sj_data_for_model %>% mutate(predicted_num_cases = sj_predicts)
iq_data_for_model <- iq_data_for_model %>% mutate(predicted_num_cases = iq_predicts)

sj_driven_data_predictions <- write.csv(select(sj_data_for_model, year, week_of_year, predicted_num_cases), "sj_driven_data_predictions")
iq_driven_data_predictions <- write.csv(select(iq_data_for_model, year, week_of_year, predicted_num_cases), "iq_driven_data_predictions")
```

